Report for Qualifying Exam

### 2024-07-22 src/sections/relatedwork.tex
added related work section, pending:table citation, caption

\begin{itemize}
    \item \textbf{Investigating Semantic Shifts Through Word Contexts.}
    Gulordava and Baroni (2011) detected semantic change by focusing on words used in the 1960s and 1990s.
        They compared the similarity of the surrounding words (words that co-occur with the target word) in these two time periods.
    \begin{itemize}
        \item To assess similarity, the researchers calculated the \emph{Local Mutual Information (LMI)} score between the central word and its surrounding words.
        A low LMI score between the target word and its surrounding words across the two time periods indicated a potential semantic shift.
        \item The study demonstrated that their distributional similarity models were effective in capturing \emph{cultural shifts} in word meaning.
        For example, they found that the word `sleep' acquired more negative connotations related to sleep disorders when comparing its contexts in the 1960s to those in the 1990s.
    \end{itemize}

@@

% --------------------------------- SYNTACTIC ---------------------------------
\para{Linguistic change studies have shifted to include \emph{syntax} alongside semantics.}
The field of linguistic change has traditionally focused on \emph{semantics}.
However, recent studies have begun to explore the role of \emph{syntax} in language evolution as well.
The \emph{syntactic} functionality of a word can evolve by transitioning into a new part-of-speech (POS) category.
Nouns, due to their inherent flexibility in meaning, exhibit a greater tendency to undergo these changes driven by cultural shifts.
While verbs are more likely to participate in gradual semantic changes that follow established linguistic patterns.
\vspace{0mm}
\begin{itemize}
    \item \para{Acquiring a new POS.}
    Kulkarni et al.\ assigned part-of-speech (POS) tags to a large collection of text.
    They calculated the likelihood of a word appearing in specific grammatical contexts over time.
    \begin{itemize}
        \item To quantify temporal change, they compared the probability distributions of POS tags for a particular word across different time periods.
        This essentially measured the divergence between these distributions.
        \item An example they cited was the word `windows'.
        Its POS tag shifted from a common noun (referring to doors and windows of a house) to a proper noun (`Microsoft Windows').
        This highlights how a word's grammatical function can change alongside its meaning.
    \end{itemize}

    \item \para{Capturing Differences Between Nouns and Verbs.}
     Hamilton et al.\ (2016) highlight how \emph{cultural changes}, often influenced by new technologies, are closely tied to transformations within \emph{local neighborhoods}, particularly sensitive to shifts in nouns.
     On the other hand, \emph{linguistic changes} are more associated with \emph{global measures} and are particularly responsive to variations in verbs.
     \begin{itemize}
         \item To validate this hypothesis, the authors employed a statistical technique called a \emph{linear mixed model} where word type (noun or verb) is a fixed effect,
         amount of change measured by each metric (local or global) treated as the dependent variable.
         By analyzing the model's results, they could assess whether there's a significant difference in the way nouns and verbs exhibit change.
         \item The evolution of words like `actually', `must', and `promise' demonstrate these changes.
         For instance, `must' has transitioned from expressing obligation to indicating necessity, showcasing a common pattern seen in modal verbs.
     \end{itemize}

%    Cultural changes like new technologies are related to local neighbourhood change, which is more sensitive to changes in nouns.
%    Examples: gay, virus, cell
%    Linguistic changes like subjectification is related to global measure, which is more sensitive to change in verbs.
%    Examples: actually, must, promise
%    actually is a case of subjectification
%    must shifted from a deontic/obligation usage (“you must do X”) to a epistemic one (“X must be the case”), exemplifying a regular pattern of change common to many modal verb
%    promise represents the class of shifting “performative speech acts” that undergo rich changes due to their pragmatic uses and subjectification

%    \item \para{Acquiring a new sense:}
%    Words are complex entities with multiple senses that can evolve over time while their form remains constant, as discussed by (Laurino et al., 2023).
%    Their research studies the overall mental lexicon of words, focusing on quantifying the drift of word concepts.
%    The study highlights how most words possess a range of meanings that can be added, removed, or modified across different temporal contexts.
\end{itemize}

\vspace{5mm}
%Among the reviewed literature, three studies are particularly pertinent to our analysis of data voids.
%In case of data voids, we want to find shifts from it's previous usage to new usage.
Several studies from our literature review provide valuable insights for addressing data voids in our analysis.
\begin{itemize}
    \item The first being the work by Kulkarni et al.\ which introduces us to time series construction.
Their distributional methods focuses on finding subtle semantic shits to determine the context where a word in used.
This concept aligns perfectly with our goal of uncovering shifts in word usage when encountering data voids.
    \item Second paper by Hamilton et al.\ introduces us to word emebeddings alignment.
Since we want to compare word vectors from different time periods, vectors should be aligned in same coordinate axes.
After aligning the embeddings for individual time periods, we can use the aligned word vectors to compute the semantic displacement that a word has undergone during a certain time-period.
    \item Third one by Laurino et al.\ investigates the impact of the COVID-19 pandemic on word meaning.
One key finding from their work is that words directly related to the pandemic exhibited a greater difference in semantic similarity between pre-pandemic and pandemic time periods.
This suggests that these words underwent a more significant and rapid semantic shift compared to control words not associated with the pandemic.
They also employ semantic similarity analysis to quantify the shifts in meaning for pandemic-related words and provides evidence that the COVID-19 pandemic acted as a catalyst for rapid semantic change.
\end{itemize}

\begin{table}[t]
\centering
\begin{tabular}{lllll}
\toprule
\textbf{Causes}    & \textbf{Methods}        & \textbf{Technical Method} & \textbf{Time Period} & \textbf{Example Papers} \\ \hline
\textbf{Semantic}  & \textbf{Frequency}      & Log ratio                 & 1960 \& 1990         & {[}1{]}                 \\ \hline
                   &                         & Word Frequencies          & 1900 - 2005          & {[}5{]}                 \\ \cline{3-5}
                   &                         & Cosine Distance           &                      &                         \\ \cline{2-5}
                   & \textbf{Distributional} & Co-occurence              & 1960 \& 1990         & {[}1{]}                 \\ \cline{2-5}
                   &                         & NLM                       &                      & {[}4{]}                 \\ \cline{3-5}
                   &                         & Cosine Similarity         & 1900 - 2005          & {[}5{]}                 \\ \cline{3-5}
                   &                         & Aligning Embeddings       &                      & {[}3{]}                 \\ \cline{3-5}
                   &                         & DBSCAN                    &                      & {[}8{]}                 \\ \cline{3-5}
                   &                         & BERT                      &                      & {[}8{]}                 \\ \hline
\textbf{Syntactic} & \textbf{POS Tags}       & JS Divergence             & 1900-2000            & {[}3{]}                 \\ \hline
                   &                         & Cosine Distance           & 1800 - 1990          & {[}2{]}                 \\ \cline{3-5}
\textbf{}          & \textbf{}               & KL Divergence             & 2020 \& 2023         & {[}7{]}                 \\
\bottomrule
\end{tabular}
\caption{Various types of linguistic change and method used to detect them.}
\label{tab:related-work}
\end{table}

>>>>

\begin{table}[]
\centering
\small
\begin{tabular}{@{}llllll@{}}
\toprule
\multicolumn{1}{c}{\textbf{\begin{tabular}[c]{@{}c@{}}Linguistic \\ Change\end{tabular}}} & \multicolumn{1}{c}{\textbf{Approach}}                                              & \multicolumn{1}{c}{\textbf{TCD}} & \multicolumn{3}{c}{\textbf{Example}}                                                                                                                                                                 \\ \cmidrule(l){4-6}
\multicolumn{1}{c}{\textbf{}}                                                             & \multicolumn{1}{c}{\textbf{}}                                              & \multicolumn{1}{c}{\textbf{}}    & \textbf{Word}         & \textbf{Former Usage}                                                                & \textbf{New Usage}                                                                    \\ \midrule
\multicolumn{1}{l}{\textbf{Semantic}}                                                     & \multicolumn{1}{l}{\textbf{FREQUENCY}}                                             & \multicolumn{1}{l}{}            & \multicolumn{1}{l}{} & \multicolumn{1}{l}{}                                                                & \multicolumn{1}{l}{}                                                                 \\ \midrule
\textbf{}                                                                                 & Log Ratio                                                                  & 1960 \& 1990                     & disk                  & -                                                                                    & -                                                                                     \\ \cmidrule(l){2-6}
\textbf{}                                                                                 & \begin{tabular}[c]{@{}l@{}}Word \\ Frequency\end{tabular}                  & 1900--2005                      & bitch                 & female dog                                                                           & slang                                                                                 \\ \cmidrule(l){2-6}
\textbf{}                                                                                 & \begin{tabular}[c]{@{}l@{}}Law of \\ Conformity\end{tabular}               & 1800--2000                      & -                     & -                                                                                    & -                                                                                     \\ \cmidrule(l){2-6}
                                                                                          & \textbf{DISTRIBUTIONAL}                                                             &                                  &                       &                                                                                      &                                                                              \\ \cmidrule(l){2-6}
\textbf{}                                                                                 & \begin{tabular}[c]{@{}l@{}}Local Mutual \\ Information\end{tabular}        & 1960 \& 1990                     & sleep                 & deep sleep                                                                           & sleep disorder                                                                        \\ \cmidrule(l){2-6}
\textbf{}                                                                                 & \begin{tabular}[c]{@{}l@{}}Continuous \\ Word \\ Embeddings\end{tabular}   & 1900--2009                      & cell                  & closet, dungeon                                                                      & phone, cordless                                                                       \\ \cmidrule(l){2-6}
\textbf{}                                                                                 & \begin{tabular}[c]{@{}l@{}}Change Point \\ Algorithm\end{tabular}          & 1900--2005                      & gay                   & cheerful, dapper                                                                     & \begin{tabular}[c]{@{}l@{}}lesbian, \\ homosexual\end{tabular}                        \\ \cmidrule(l){2-6}
\textbf{}                                                                                 & Clustering (DBSCAN)                                                        & 1900--2000                      & mouse                 & mice, rat                                                                            & cursor, pointer                                                                       \\ \cmidrule(l){2-6}
\textbf{}                                                                                 & \begin{tabular}[c]{@{}l@{}}Cultural and \\ linguistic Drift\end{tabular}   & 1800--2000                      & virus                 & \begin{tabular}[c]{@{}l@{}}infected with\\  the virus\end{tabular}                   & \begin{tabular}[c]{@{}l@{}}spreading\\  computer virus\end{tabular}                   \\ \cmidrule(l){2-6}
\textbf{}                                                                                 & \begin{tabular}[c]{@{}l@{}}Diachronic \\ Word\\ Embeddings\end{tabular}    & 1800--1999                      & awful                 & full of awe                                                                          & terrible or appalling                                                                 \\ \cmidrule(l){2-6}
\textbf{}                                                                                 & \begin{tabular}[c]{@{}l@{}}Contextual \\ Word \\ Embeddings\end{tabular}                                                                      & 1910 – 2009                      & tenure                & \begin{tabular}[c]{@{}l@{}}short term leases,\\ insecurity of tenure\end{tabular}    & \begin{tabular}[c]{@{}l@{}}tenure of office, \\ employment \\ and tenure\end{tabular} \\ \cmidrule(l){2-6}
\textbf{}                                                                                 & \begin{tabular}[c]{@{}l@{}}Semantic \\ Similarity \\ Analysis\end{tabular} & 2014--2022                      & immunity              & \begin{tabular}[c]{@{}l@{}}politics (‘legislator’,\\ ‘representative’).\end{tabular} & \begin{tabular}[c]{@{}l@{}}health-related \\ (‘prevention’, ‘AIDS’)\end{tabular}      \\ \cmidrule(l){2-6}
                                                                                          & \textbf{PART OF SPEECH}                                                                        &                                  &                       &                                                                                      &                                                                   \\ \midrule
\textbf{Syntactic}                                                                        & POS Tags                                                                   & 1900--2000                        & windows               & \begin{tabular}[c]{@{}l@{}}doors and \\ windows of a house\end{tabular}              & Microsoft Windows                                                                     \\ \midrule
\textbf{}                                                                                 & \begin{tabular}[c]{@{}l@{}}Mixed-model \\ Regressions\end{tabular}         & 1800--2000                      & actually              & originally, nominally                                                                & presumed, believe\\

\bottomrule
\end{tabular}
%\bottomrule
\caption{Various types of linguistic change and method used to detect them.}
\label{tab:related-work}
\end{table}

% ------------------------------------------------------------------ SEMANTIC ------------------------------------------------------------------
\begin{itemize}
    \item \textbf{Investigating Semantic Shifts Through Word Contexts.}
    Gulordava and Baroni (2011) detected semantic change by focusing on words used in the 1960s and 1990s.
        They compared the similarity of the surrounding words (words that co-occur with the target word) in these two time periods.
    \begin{itemize}
        \item To assess similarity, the researchers calculated the \emph{Local Mutual Information (LMI)} score between the central word and its surrounding words.
        A low LMI score between the target word and its surrounding words across the two time periods indicated a potential semantic shift.
        \item The study demonstrated that their distributional similarity models were effective in capturing \emph{cultural shifts} in word meaning.
        For example, they found that the word `sleep' acquired more negative connotations related to sleep disorders when comparing its contexts in the 1960s to those in the 1990s.
    \end{itemize}

@@

% ------------------------------------------------------------------ SYNTACTIC ------------------------------------------------------------------
\para{Linguistic change studies have shifted to include \emph{syntax} alongside semantics.}
The field of linguistic change has traditionally focused on \emph{semantics}.
However, recent studies have begun to explore the role of \emph{syntax} in language evolution as well.
The \emph{syntactic} functionality of a word can evolve by transitioning into a new part-of-speech (POS) category.
Nouns, due to their inherent flexibility in meaning, exhibit a greater tendency to undergo these changes driven by cultural shifts.
While verbs are more likely to participate in gradual semantic changes that follow established linguistic patterns.
\vspace{0mm}
\begin{itemize}
    \item \para{Acquiring a new POS.}
    Kulkarni et al.\ assigned part-of-speech (POS) tags to a large collection of text.
    They calculated the likelihood of a word appearing in specific grammatical contexts over time.
    \begin{itemize}
        \item To quantify temporal change, they compared the probability distributions of POS tags for a particular word across different time periods.
        This essentially measured the divergence between these distributions.
        \item An example they cited was the word `windows'.
        Its POS tag shifted from a common noun (referring to doors and windows of a house) to a proper noun (`Microsoft Windows').
        This highlights how a word's grammatical function can change alongside its meaning.
    \end{itemize}

    \item \para{Capturing Differences Between Nouns and Verbs.}
     Hamilton et al.\ (2016) highlight how \emph{cultural changes}, often influenced by new technologies, are closely tied to transformations within \emph{local neighborhoods}, particularly sensitive to shifts in nouns.
     On the other hand, \emph{linguistic changes} are more associated with \emph{global measures} and are particularly responsive to variations in verbs.
     \begin{itemize}
         \item To validate this hypothesis, the authors employed a statistical technique called a \emph{linear mixed model} where word type (noun or verb) is a fixed effect,
         amount of change measured by each metric (local or global) treated as the dependent variable.
         By analyzing the model's results, they could assess whether there's a significant difference in the way nouns and verbs exhibit change.
         \item The evolution of words like `actually', `must', and `promise' demonstrate these changes.
         For instance, `must' has transitioned from expressing obligation to indicating necessity, showcasing a common pattern seen in modal verbs.
     \end{itemize}
\end{itemize}

% ------------------------------------------------------------------ THREE PAPERS ------------------------------------------------------------------
Several studies from our literature review provide valuable insights for addressing data voids in our analysis.
\begin{itemize}
    \item The first being the work by Kulkarni et al.\ which introduces us to time series construction.
Their distributional methods focuses on finding subtle semantic shits to determine the context where a word in used.
This concept aligns perfectly with our goal of uncovering shifts in word usage when encountering data voids.
    \item Second paper by Hamilton et al.\ introduces us to word emebeddings alignment.
Since we want to compare word vectors from different time periods, vectors should be aligned in same coordinate axes.
After aligning the embeddings for individual time periods, we can use the aligned word vectors to compute the semantic displacement that a word has undergone during a certain time-period.
    \item Third one by Laurino et al.\ investigates the impact of the COVID-19 pandemic on word meaning.
One key finding from their work is that words directly related to the pandemic exhibited a greater difference in semantic similarity between pre-pandemic and pandemic time periods.
This suggests that these words underwent a more significant and rapid semantic shift compared to control words not associated with the pandemic.
They also employ semantic similarity analysis to quantify the shifts in meaning for pandemic-related words and provides evidence that the COVID-19 pandemic acted as a catalyst for rapid semantic change.
\end{itemize}
